Revisiting the Relationship between Short-Chain Fatty Acids, the Microbiota, and Colorectal Tumors

### 2019-04-03 submission/manuscript.Rmd
Render manuscript

# Fecal short chain fatty acids are not predictive of colorectal cancer status and cannot be predicted based on bacterial community structure

@@

Marc A. Sze${^1}$, Begüm Topçuoğlu${^1}$, Nicholas A. Lesniak${^1}$, Mack T. Ruffin IV${^2}$, Patrick D. Schloss${^1}$${^\dagger}$

@@

Colonic bacterial populations are thought to have a role in the development of colorectal cancer with some protecting against inflammation and others exacerbating inflammation. Short chain fatty acids (SCFAs), including butyrate, have been shown to have anti-inflammatory properties and are produced in large quantities by colonic bacteria which produce SCFAs by fermenting fiber. We assessed whether there was an association between fecal SCFA concentrations and the presence of colonic adenomas or carcinomas in a cohort of individuals using 16S rRNA gene and metagenomic shotgun sequence data. We measured the fecal concentrations of acetate, propionate, and butyrate within the cohort and found that there were no meaningful association between SCFA concentration and disease status. When we incorporated these concentrations into Random Forest classification models trained to differentiate between people with normal colons and those with adenomas or carcinomas, we found that they did not significantly improve the ability of 16S rRNA gene or metagenomic gene sequence-based models to classify individuals. Finally, we generated Random Forest regression models trained to predict the concentration of each SCFA based on 16S rRNA gene or metagenomic gene sequence data from the same samples. These models performed poorly and were able to explain at most 14% of the observed variation in the SCFA concentrations. These results support the broader epidemiological data that questions the value of fiber consumption for reducing the risks of colorectal cancer. Although other bacterial metabolites may serve as biomarkers to detect adenomas or carcinomas, fecal SCFA concentrations have limited predictive power.

@@

**SCFA concentrations do not meaningfully vary with diagnosis or treatment.** To quantify the associations between colorectal cancer, the microbiome, and SCFAs, we quantified the concentration of acetate, propionate, and butyrate in feces of previously characterized individuals with normal colons (N=`r n_normal`) and those with colonic adenomas (N=`r n_adenoma`) or carcinomas (N=`r n_cancer`) [@Baxter2016]. We were unable to detect a significant difference in any SCFA concentration across the diagnoses groups (all P>`r p_cross_min`; Figure 1A). Among the individuals with adenomas and carcinomas, a subset (N~adenoma~=`r n_follow_adenoma`, N~carcinoma~=`r n_follow_cancer`) were treated and sampled a year later [@Sze2017]. None of the SCFAs exhibited a significant change with treatment (all P>`r p_follow_min`; Figure 1B). For both the pre-treatment cross-sectional data and the pre/post treatment data, we also failed to detect any significant differences in the relative concentrations of any SCFAs (P>`r rel_test`). Finally, we pooled the SCFA concentrations on a total and per molecule of carbon basis and again failed to observe any significant differences (P>`r pooled_test`). These results demonstrated that there were no significant associations between fecal SCFA concentration and diagnosis or treatment.

@@

**Combining SCFA and microbiome data does not improve the ability to diagnose individual as having adenomas or carcinomas.** We previously found that binning 16S rRNA gene sequence data into operational taxonomic units (OTUs) based on 97% similarity or into genera enabled us to classify individuals as having adenomas or carcinomas using Random Forest machine learning models [@Baxter2016; @Sze2018]. We repeated that analysis but added the concentration of the SCFAs as possible features to train the models (Figure S1). Models trained using SCFAs to classify individuals as having adenomas or carcinomas rather than normal colons had median areas under the receiver operator characteristic curve (AUROC) that were significantly greater than 0.5 (P~adenoma~<0.001 and P~carcinoma~<0.001); however, the AUROC values to detect the presence of adenomas or carcinomas were only `r random %>% filter(class=="adenoma") %>% pull(median)` and `r random %>% filter(class=="cancer") %>% pull(median)`, respectively (Figure 2A). When we trained the models with the SCFAs concentrations and OTU or genus-level relative abundances the AUROC values were not significantly different from the same models trained without the SCFA concentrations (P>`r min_p_compare_tax`; Figure 2A). These data demonstrate that knowledge of the SCFA profile from a patient's fecal sample did not improve the ability to diagnose a colonic lesion.

@@

**Knowledge of microbial community structure does not predict SCFA concentrations.** We next asked whether the fecal community structure was predictive of fecal SCFA concentrations, regardless of a person's diagnosis. We trained Random Forest regression models using 16S rRNA gene sequence data binned into OTUs and genera to predict the concentration of the SCFAs (Figure S2). The largest R^2^ between the observed SCFA concentrations and the modeled concentrations was `r top_16S$median_Rsquared`, which was observed when using `r convert[top_16S$microbiome]` data to predict `r top_16S$class` concentrations (Figure 2B). We also used a smaller dataset of shotgun metagenomic sequencing data generated from a subset of our cohort (N~normal~=`r n_s %>% filter(dx == "normal") %>% pull(n)`, N~adenoma~=`r n_s %>% filter(dx == "adenoma") %>% pull(n)`, and N~cancer~=`r n_s %>% filter(dx == "cancer") %>% pull(n)`) [@Hannigan2017]. We binned genes extracted from the assembled metagenomes into operational protein families (OPFs) or KEGG categories and trained Random Forest regression models using metagenomic sequence data to predict the concentration of the SCFAs (Figure S2). Similar to the analysis using 16S rRNA gene sequence data, the metagenomic data was not predictive of SCFA concentration; the largest amount of variation that the models could explain was `r top_metagenomics$median_Rsquared`, which was observed when using `r convert[top_metagenomics$microbiome]` data to predict `r top_metagenomics$class` concentrations (Figure 2B). Because of the limited number of samples that we were able to generate metagenomic sequence data from, we used our 16S rRNA gene sequence data to impute metagenomes that were binned into metabolic pathways or KEGG categories using PICRUSt (Figure S2). SCFA concentrations could not be predicted based on the imputed metagenomic data. The largest amount of variation that the models could explain was `r top_picrust$median_Rsquared`, which was observed when using `r convert[top_picrust$microbiome]` data to predict `r top_picrust$class` concentrations (Figure 2B). The inability to model SCFA concentrations from microbiome data indicates that the knowledge of the abundance of organisms and their genes was insufficient to predict SCFA concentrations.

@@

**Random Forest models.** The classification models were built to predict lesion type from microbiome information with or without SCFA concentrations. The regression models were built to predict the SCFA concentrations of acetate, butyrate, and propionate from microbiome information. For classification and regression models, we pre-processed the features by scaling them to vary between zero and one. Features with no variance in the training set were removed from both the training and testing sets. We randomly split the data into training and test sets so that the training set consisted of 80% of the full dataset while the test set was composed of the remaining data. The training set was used for hyperparameter selection and training the model and the test set was used for evaluating prediction performance. For each model, the best performing hyperparameter, mtry, was selected in an internal five-fold cross-validation of the training set with 100 randomizations. Six values of mtry were tested and the value that provided the largest AUROC or R^2^ was selected. We trained the random forest model using the selected mtry value and predicted the held-out test set. The data-split, hyperparameter selection, training and testing steps were repeated 100 times to get a reliable and robust reading of model prediction performance. We used AUROC and R^2^ as the prediction performance metric for classification and regression models, respectively. We used randomForest package implemented to the caret package (version `r installed.packages()["randomForest" ,"Version"]`) in R statistical software (version `r installed.packages()["caret" ,"Version"]`) for our models.

**Statistical analysis workflow.** Data summaries, statistical analysis, and data visualizations were performed using R (v.3.5.1) with the tidyverse package (v.1.2.1). To assess differences in SCFA concentrations between individuals normal colons and those with adenomas or carcinomas, we used the Kruskal-Wallis rank sum test. If a test had a P-value below 0.05, we then applied a pairwise Wilcoxon rank sum test with a Benjamini-Hochberg correction for multiple comparisons. To assess differences in SCFA concentrations between individuals samples before and after treatment we used paired Wilcoxon rank sum tests to test for significance. To compare the median AUCROC for the held out data for the model generated using only the SCFAs, we compared the distribution of the data to the expected median of 0.5 using the Wilcoxon rank sum test to test whether the model performed better than would be achieved by randomly assigning the data to each diagnosis. When we compared the Random Forest models generated without and with SCFA data included, we used Wilcoxon rank sum tests to determine whether the models with the SCFA data included did better.

@@

**Figure 2. SCFA concentrations do not improve models for diagnosing the presence of adenomas, carcinomas, or all lesions and cannot be reliably predicted from 16S rRNA gene or metagenomic sequence data.** (A) The median AUROC for diagnosing individuals as having adenomas or carcinomas using SCFAs was slightly better than than chance (depicted by horizontal line at 0.50), but did improve performance of the models generated using 16S rRNA gene sequence data. (B) Regression models that were trained using 16S rRNA gene sequence, metagenomic, and PICRUSt data to predict the concentrations of SCFAs performed poorly (all median R^2^ values < `r max(regression$median_Rsquared)`). Regression models generated using 16S rRNA gene sequence and PICRUSt data included data from `r sum(n$n)` samples and those generated using metagenomic data included data from `r sum(n_s$n)` samples.

@@

**Figure S1. Comparison of training and testing results for classification models shows that the models are robust and are not overfit.** Random Forest classification models were generated to differentiate between individuals with normal colons and those with adenomas or carcinomas using 16S rRNA gene sequence data that were clustered into genera or OTUs with and without including the three SCFAs as additional features. Random Forest classification models were generated by partitioning the samples into a training set with 80% of the data and a testing set with the remaining samples for 100 randomizations.

@@

**Figure S2. Comparison of training and testing results for regression models shows that the models are robust and are not overfit.** Random Forest regression models were generated to predict the concentration of each SCFA using each individuals' microbiome data generated using 16S rRNA gene sequence and metagenomic sequence data. These regression models were generated by partitioning the samples into a training set with 80% of the data and a testing set with the remaining samples for 100 randomizations.

>>>>

# Fecal short chain fatty acids are not predictive of colonic tumor status and cannot be predicted based on bacterial community structure

@@

Marc A. Sze${^1}$, Begüm D. Topçuoğlu${^1}$, Nicholas A. Lesniak${^1}$, Mack T. Ruffin IV${^2}$, Patrick D. Schloss${^1}$${^\dagger}$

@@

Colonic bacterial populations are thought to have a role in the development of colorectal cancer with some protecting against inflammation and others exacerbating inflammation. Short chain fatty acids (SCFAs), including butyrate, have been shown to have anti-inflammatory properties and are produced in large quantities by colonic bacteria which produce SCFAs by fermenting fiber. We assessed whether there was an association between fecal SCFA concentrations and the presence of colonic adenomas or carcinomas in a cohort of individuals using 16S rRNA gene and metagenomic shotgun sequence data. We measured the fecal concentrations of acetate, propionate, and butyrate within the cohort and found that there were no significant associations between SCFA concentration and disease status. When we incorporated these concentrations into random forest classification models trained to differentiate between people with normal colons and those with adenomas or carcinomas, we found that they did not significantly improve the ability of 16S rRNA gene or metagenomic gene sequence-based models to classify individuals. Finally, we generated random forest regression models trained to predict the concentration of each SCFA based on 16S rRNA gene or metagenomic gene sequence data from the same samples. These models performed poorly and were able to explain at most 14% of the observed variation in the SCFA concentrations. These results support the broader epidemiological data that questions the value of fiber consumption for reducing the risks of colorectal cancer. Although other bacterial metabolites may serve as biomarkers to detect adenomas or carcinomas, fecal SCFA concentrations have limited predictive power.

@@

**Fecal SCFA concentrations did not vary with diagnosis or treatment.** To quantify the associations between colorectal cancer, the microbiome, and SCFAs, we quantified the concentration of acetate, propionate, and butyrate in feces of previously characterized individuals with normal colons (N=`r n_normal`) and those with colonic adenomas (N=`r n_adenoma`) or carcinomas (N=`r n_cancer`) [@Baxter2016]. We were unable to detect a significant difference in any SCFA concentration across the diagnoses groups (all P>`r p_cross_min`; Figure 1A). Among the individuals with adenomas and carcinomas, a subset (N~adenoma~=`r n_follow_adenoma`, N~carcinoma~=`r n_follow_cancer`) were treated and sampled a year later [@Sze2017]. None of the SCFAs exhibited a significant change with treatment (all P>`r p_follow_min`; Figure 1B). For both the pre-treatment cross-sectional data and the pre/post treatment data, we also failed to detect any significant differences in the relative concentrations of any SCFAs (P>`r rel_test`). Finally, we pooled the SCFA concentrations on a total and per molecule of carbon basis and again failed to observe any significant differences (P>`r pooled_test`). These results demonstrated that there were no significant associations between fecal SCFA concentration and diagnosis or treatment.

@@

**Combining SCFA and microbiome data does not improve the ability to diagnose individual as having adenomas or carcinomas.** We previously found that binning 16S rRNA gene sequence data into operational taxonomic units (OTUs) based on 97% similarity or into genera enabled us to classify individuals as having adenomas or carcinomas using random forest machine learning models [@Baxter2016; @Sze2018]. We repeated that analysis but added the concentration of the SCFAs as possible features to train the models (Figure S1). Models trained using SCFAs to classify individuals as having adenomas or carcinomas rather than normal colons had median areas under the receiver operator characteristic curve (AUROC) that were significantly greater than 0.5 (P~adenoma~<0.001 and P~carcinoma~<0.001). However, the AUROC values to detect the presence of adenomas or carcinomas were only `r random %>% filter(class=="adenoma") %>% pull(median)` and `r random %>% filter(class=="cancer") %>% pull(median)`, respectively, indicating that SCFAs had poor predictive power on their own (Figure 2A). When we trained the models with the SCFAs concentrations and OTU or genus-level relative abundances the AUROC values were not significantly different from the same models trained without the SCFA concentrations (P>`r min_p_compare_tax`; Figure 2A). These data demonstrate that knowledge of the SCFA profile from a patient's fecal sample did not improve the ability to diagnose a colonic lesion.

@@

**Knowledge of microbial community structure does not predict SCFA concentrations.** We next asked whether the fecal community structure was predictive of fecal SCFA concentrations, regardless of a person's diagnosis. We trained random forest regression models using 16S rRNA gene sequence data binned into OTUs and genera to predict the concentration of the SCFAs (Figure S2). The largest R^2^ between the observed SCFA concentrations and the modeled concentrations was `r top_16S$median_Rsquared`, which was observed when using `r convert[top_16S$microbiome]` data to predict `r top_16S$class` concentrations (Figure 2B). We also used a smaller dataset of shotgun metagenomic sequencing data generated from a subset of our cohort (N~normal~=`r n_s %>% filter(dx == "normal") %>% pull(n)`, N~adenoma~=`r n_s %>% filter(dx == "adenoma") %>% pull(n)`, and N~cancer~=`r n_s %>% filter(dx == "cancer") %>% pull(n)`) [@Hannigan2017]. We binned genes extracted from the assembled metagenomes into operational protein families (OPFs) or KEGG categories and trained random forest regression models using metagenomic sequence data to predict the concentration of the SCFAs (Figure S2). Similar to the analysis using 16S rRNA gene sequence data, the metagenomic data was not predictive of SCFA concentration; the largest amount of variation that the models could explain was `r 100*top_metagenomics$median_Rsquared`%, which was observed when using `r convert[top_metagenomics$microbiome]` data to predict `r top_metagenomics$class` concentrations (Figure 2B). Because of the limited number of samples that we were able to generate metagenomic sequence data from, we used our 16S rRNA gene sequence data to impute metagenomes that were binned into metabolic pathways or KEGG categories using PICRUSt (Figure S2). SCFA concentrations could not be predicted based on the imputed metagenomic data. The largest amount of variation that the models could explain was `r 100*top_picrust$median_Rsquared`%, which was observed when using `r convert[top_picrust$microbiome]` data to predict `r top_picrust$class` concentrations (Figure 2B). The inability to model SCFA concentrations from microbiome data indicates that the knowledge of the abundance of organisms and their genes was insufficient to predict SCFA concentrations.

@@

**random forest models.** The classification models were built to predict lesion type from microbiome information with or without SCFA concentrations. The regression models were built to predict the SCFA concentrations of acetate, butyrate, and propionate from microbiome information. For classification and regression models, we pre-processed the features by scaling them to vary between zero and one. Features with no variance in the training set were removed from both the training and testing sets. We randomly split the data into training and test sets so that the training set consisted of 80% of the full dataset while the test set was composed of the remaining data. The training set was used for hyperparameter selection and training the model and the test set was used for evaluating prediction performance. For each model, the best performing hyperparameter, mtry, was selected in an internal five-fold cross-validation of the training set with 100 randomizations. Six values of mtry were tested and the value that provided the largest AUROC or R^2^ was selected. We trained the random forest model using the selected mtry value and predicted the held-out test set. The data-split, hyperparameter selection, training and testing steps were repeated 100 times to get a reliable and robust reading of model prediction performance. We used AUROC and R^2^ as the prediction performance metric for classification and regression models, respectively. We used randomForest package implemented to the caret package (version `r installed.packages()["randomForest" ,"Version"]`) in R statistical software (version `r installed.packages()["caret" ,"Version"]`) for our models.

**Statistical analysis workflow.** Data summaries, statistical analysis, and data visualizations were performed using R (v.3.5.1) with the tidyverse package (v.1.2.1). To assess differences in SCFA concentrations between individuals normal colons and those with adenomas or carcinomas, we used the Kruskal-Wallis rank sum test. If a test had a P-value below 0.05, we then applied a pairwise Wilcoxon rank sum test with a Benjamini-Hochberg correction for multiple comparisons. To assess differences in SCFA concentrations between individuals samples before and after treatment we used paired Wilcoxon rank sum tests to test for significance. To compare the median AUCROC for the held out data for the model generated using only the SCFAs, we compared the distribution of the data to the expected median of 0.5 using the Wilcoxon rank sum test to test whether the model performed better than would be achieved by randomly assigning the data to each diagnosis. When we compared the random forest models generated without and with SCFA data included, we used Wilcoxon rank sum tests to determine whether the models with the SCFA data included did better.

@@

**Figure 2. SCFA concentrations do not improve models for diagnosing the presence of adenomas, carcinomas, or all lesions and cannot be reliably predicted from 16S rRNA gene or metagenomic sequence data.** (A) The median AUROC for diagnosing individuals as having adenomas or carcinomas using SCFAs was slightly better than than chance (depicted by horizontal line at 0.50), but did not improve performance of the models generated using 16S rRNA gene sequence data. (B) Regression models that were trained using 16S rRNA gene sequence, metagenomic, and PICRUSt data to predict the concentrations of SCFAs performed poorly (all median R^2^ values < `r max(regression$median_Rsquared)`). Regression models generated using 16S rRNA gene sequence and PICRUSt data included data from `r sum(n$n)` samples and those generated using metagenomic data included data from `r sum(n_s$n)` samples.

@@

**Figure S1. Comparison of training and testing results for classification models shows that the models are robust and are not overfit.** random forest classification models were generated to differentiate between individuals with normal colons and those with adenomas or carcinomas using 16S rRNA gene sequence data that were clustered into genera or OTUs with and without including the three SCFAs as additional features. random forest classification models were generated by partitioning the samples into a training set with 80% of the data and a testing set with the remaining samples for 100 randomizations.

@@

**Figure S2. Comparison of training and testing results for regression models shows that the models are robust and are not overfit.** random forest regression models were generated to predict the concentration of each SCFA using each individuals' microbiome data generated using 16S rRNA gene sequence and metagenomic sequence data. These regression models were generated by partitioning the samples into a training set with 80% of the data and a testing set with the remaining samples for 100 randomizations.
